Learning with misspecified models: Overconfidence and Stereotypes"  # nolint: Grammarly.

### 2023-10-06 Paper/draft.Rmd
learners and not learners

# Related Literature

# Framework 1

An agent is of type $\theta \in \Theta$ and faces an unknown exogenous state $\omega$ drawn from some 
density $f$ over $\Omega$. The agent knows the 
distribution of $\omega$ but not its realized value. His prior belief about his type is $p_0(\theta)$ and his belief about the state
state $p_0(\omega)$ coincides with the true distribution $f$. Let the agent's type be $\theta^*$ and the realized state be $\omega^*$.

An agent has a \emph{misspecified} belief if the prior assigns probability zero to their true type. Furthermore, the agent is 
\emph{dogmatic} if he holds a degenerate belief that places probability one on being a particular type, $\hat{\theta}$. An 
agent can be dogmatic and misspecified; that means that $\hat{\theta} \neq \theta^*$ and $p_0(\hat{\theta}) = 1$.

@@

After observing the outcome, the agent updates his beliefs about $\theta$ and $\omega$ using some algorithm and moves on to the next period.
He repeats this process infinitely many times. I make the simplifying assumption that the agent is myopic and chooses the action 
that maximizes the payoff in each period. This assumption is symplifies the analysis and plays a role on whether an agent who updates
 their beliefs using bayes rule would learn the truth or not.

A key notion in this setting is that of a self-defeating equilibrium\footnote{This notion is an adaptation of the Berk-Nash equilibrium in 
@Esponda2016 to this setting with only one agent}. 
A \emph{self-defeating equilibrium} is a belief and action pair such that the agent's belief about their type is misspecified and the 
outcome generated by the action is consistent with the misspecified belief. The average outcome under the true type and the true 
state equals the average output the agent expects under the misspecified belief. In addition, the agent's belief is said to be 
\emph{stable} when this happens.

Within this framework, I consider two nested theories of belief updating. The first one is a dogmatic modeler from @Heidhues2018.
The second one is a switcher, as in @Ba2023. The dogmatic modeler can be seen as a switcher with an infinitely sticky initial belief
and this is the sense in which the two are nested. Both theories produce different predictions with respect to the equilibrium outcome.

## The Dogmatic Modeler

A dogmatic agent does not update their beliefs about $\theta$; instead he holds a degenerate belief that places probability one 
on being a particular type, $\hat{\theta}$, which is potentially not his true type. 
In this case, no matter
how much information he gathers against being of type $\hat{\theta}$, he will not update his beliefs. Any discrepancies between the 
observed outcomes and his believed type are incorporated using the Bayes rule to update their beliefs about $\omega$. 
This means the dogmatic agent will never learn their true type if they startu off as being misspecified.

@Heidhues2018 show that, under certain assumptions on the per-period utility,\footnote{The assumptions are that 
$u$ is twice continuously differentiable with: (i)$u_{ee}<0$ and $u_e(\underline{e} \theta, \omega)>0>u_e(\Bar{e}, \theta, \omega)$, 
(ii) $u_{\theta}, u_{\omega}>0$ and (iii) $u_{e\theta}<0$ and $u_{e\omega}>0$. The direction of the derivatives is a normalization
and the results would hold even when the signs are reversed.} 
a dogmatic modeler will inevitably fall into a self-defeating equilibrium. The equilibrium will be such that
the outcomes they observe reinforce their belief on $\omega$ in such a way that as $t\to\infty$ the agent will be sure that
the state is some $\omega^{'}$ consistent with their believed type and the observed data. In other words, 
they will be in a self-defeating equilibrium whith a stable belief that places probability one on the incorrect
parameters $(\hat{\theta}, \omega_{\infty})$

The mechanism by which the dogmatic agent falls into the self-defeating equilibrium is the following: Suppose the agent holds the 
misspecified belief that they are type $\hat{\theta}>\theta^*$. For any prior over $\omega$, the agent will be disappointed by the outcome.
He expected a gain of $h(\hat{\theta}, \mathbb{E}(\omega), a)$ but instead observes  $h(\hat{\theta}, \omega^*, a)$. 
There are two 
possible sources for the disappointment, the first is that the realized state is lower than the expected state. The second source is that the
agent is of type $\theta^*$ and therefore, for all possible states, his gain will be lower than what he expected. Because the agent is dogmatic,
he will not update his beliefs about $\theta$ and as a consequence will attribute the disappointment to the state being lower than expected.
He will continue to update in ths way until he converges to a belief about $\omega$ that is stable. Such a belief will 
explain the observed utility perfectly and allow the agent to rationalize his dogmatic belief about $\theta$. 
Under the assumptions of @Heidhues2018, there is a unique value of $\omega$ at which the belief is stable, I will refer to such value
as $\omega_\infty$.
This mechanism is further illustrated in Example 1.

@@

_Under this specification, the myopic optimal effort level is $a_t^* = \omega^*$. Nonetheless, because the agent does not know $\omega^*$,
he will choose $a_t = \mathbb{E}_t(\omega)$ where the expectation is taken with respect to the agent's belief at the beginning of
period $t$. If he does not revise his effort choice for $k$ periods, he will receive 
an average utility of 
$(\theta^*+a_t^*)\omega^* - \frac{1}{2}a_t^{*2}$ but he was expecting an average utility of 
$(\hat{\theta}\theta+a_t^*)\omega^* - \frac{1}{2}a_t^{*2}$. In response, he will apply bayes rule to update his beliefs about
$\omega$ to get the posterior belief with $\mathbb{E}_{t+k}[\omega] = \frac{(\theta^{*} + \omega^{*})\omega^{*}}{\hat{\theta} + \omega^{*}}$
which is lower than the initial belief. This will cause the agent to choose a lower effort at $t+k$. As a result, he will again
receive an average utlitity that is lower than what he expected which will cause his belief to drift further down. This process will
continue until the average utility equals his expected utility under the dogmatic belief that assigns probability 1 to $\hat{\theta}$. 
At that point, the stundent will have reached a self-defeating equilibrium and he will continue to choose sub-optimal effort forever. _

Although the model of a dogmatic modeler is able to rationalize the prevalence of overconfident (underconfident) beliefs, the
assumption that the agent has a degenerate belief and no mechanism through which he can update such belief is very restrictive. An
alternative approach is proposed by @Ba2023. She proposes an extension of the dogmatic agent who is able to jump from one dogmatic 
belief to another. By doing so, the agent might end up being dogmatic and correctly speciefied.

## The Switcher

An agent is a \emph{switcher} if they behave as a dogmatic, but is willing to entertain the possibility that they are of a different type.
In particular, when they start off as a misspecified dogmatic, they are willing to switch to a different dogmatic belief if the data is
convincing enough. Notice that their beliefs are still degenerate and assign probability one to a particular type, and zero to 
all other types. This means that a bayesian update on $\theta$ would cause no change in their beliefs.
However, they are willing to entertain two such beliefs and have a mechanism by which they decide which belief to adopt at any period $t$.

In order to abandon their initial dogmatic belief, the agent needs to observe a sequence of outcomes that are sufficiently unlikely
to have happened if they were of the type they initially believed. They do so by keeping track of the likelihood that each of the
possible types generated the data. If the likelihood ratio is sufficiently large, the agent will switch to the alternative
and behave as if they are dogmatic about the new type.

In particular, for an agent that starts off with a dogmatic belief that they are of type $\hat{\theta}$ but is willing to consider
the alternative explanation that they are of type $\tilde{\theta}$, the agent will switch to the alternative if:

@@

Where $h^t$ is the history of outcomes up to time $t$ and $\alpha$ is the switching threshold. By keeping track of the 
likelihood ratio, the agent can perform a \emph{Bayesian hypothesis test} and adopt the Dogmatic belief that best fits 
the data.\footnote{In a related problem @Schwarstein2021 proposes a similar updating procedure wich relies on the Bayesian 
hypothesis test.}
Notice that if $\alpha \to \infty$, the behavior of the switcher will be indistinguishable from that of the Dogmatic modeler. 
In this sense, the switcher is a generalization of the dogmatic type.

By allowing the agent to keep track of the likelihoods and switching to an alternative type, the switcher can avoid the 
self-confirming equilibrium. However, if the prior belief on $omega$ is sufficiently tight around
a self-defeating equilibrium, the switcher might look identical to the dogmatic even in a case where $\alpha$ is not too large.
This happens because under the agent's prior, the likelihood ratio is unlikely to grow as fast as it is needed to escape 
the self-defeating equilibrium. In this cases we say that the misspecified belief is persistent. Hence, in order to be able 
to determine which of these two models provides a better explanation of the observed behaviors, we must focus on cases 
in which the prior is diffused enough and the sequence of realized outcomes is such that
the self-defeating equilibrium would not persist.

# Framework 2

The agent is still of some type $\theta^* \in \Theta$ and the state is $\omega \sim F$. In this case the agent chooses an action 
$a\in A$
and observes a binary outcome that is either a success or a failure. Denote the outcome by $o \in {s,f}$. The probability of 
observing a success is ingreasing in $\theta^*$ and in $\omega$. Whenever the agent observes a succes, he gets a payoff $v>0$ and
whenever the outcome is a failure, normalize the payoff to 0.  In addition, the probability of success is such that for each state, 
there is a unique optimal action that maximizes the agent's expected payoff.

Two nested theories that have been widely studied fall within this framework: Fully Bayesian updating and self-serving attribution 
bias. I explain each of these classical models of belief updating in what follows

## The Bayesian

@@

## The Self-Serving Updater

@@

# A Unifying Framework
The agent can be of one of 3 types: $\theta \in \{\theta_L, \theta_M, \theta_H\}$  with $\theta_H > \theta_M > \theta_L$.
They face an unknown exogenous success rate
$\omega \in \{\omega_L, \omega_M, \omega_H\}$ with $\omega_H>\omega_M>\omega_L$. Each of the values of $\omega$ is realized with 
equal probability. 
The agent knows the distribution of $\omega$ but not its realized value.

>>>>

## Related literature

# Theoretical Framework

I consider multiple theories of belief updating that have been proposed in the literature and that are able to rationalize
the prevalence of misspecified beliefs. I focus on two different frameworks and develop a unifying example that allows me to
compare the predictions of all the theories. In this section, I outline the theories within each of the frameworks.

## Framework 1

An agent is of type $\theta \in \Theta$ and faces an unknown exogenous state $\omega$ drawn from some 
density $f$ over $\Omega$. The agent knows the 
distribution of $\omega$ but not its realized value. His prior belief about his type is $p_0(\theta)$ and his belief about the state, 
$p_0(\omega)$, coincides with the true distribution. Let the agent's true type be $\theta^{*}$,  and the realized state be $\omega^{*}$.

An agent has a \emph{misspecified} belief if the prior assigns probability zero to their true type. Furthermore, the agent is 
\emph{dogmatic} if he holds a degenerate belief that places probability one on being a particular type, $\hat{\theta}$. An 
agent can be dogmatic and misspecified. If the agent is both, $\hat{\theta} \neq \theta^*$ and $p_0(\hat{\theta}) = 1$.

@@

After observing the outcome, the agent updates his beliefs about $\theta$ and $\omega$ using some algorithm and moves on to the next period.
He repeats this process infinitely many times. I make the simplifying assumption that the agent is myopic and chooses the action 
that maximizes the payoff in each period. This assumption simplifies the analysis and plays a role in whether an agent who updates
 their beliefs using Bayes rule would learn the truth or not. The behavior of a forward-looking agent is discussed in the conclusion.

A key notion in this setting is that of a self-defeating equilibrium\footnote{This notion is an adaptation of the Berk-Nash equilibrium in 
@Esponda2016 to this setting with only one agent}. 
A \emph{self-defeating equilibrium} is a belief and action pair such that the agent's belief about their type is misspecified and the 
outcome generated by the action is consistent with the misspecified belief. This means that the average outcome under the true type and the true 
state equals the average output the agent expects under the misspecified belief. The agent's belief is said to be 
\emph{stable} when this happens.

Within this framework, I consider two nested theories of belief updating: the first one is a dogmatic modeler from @Heidhues2018,
The second one is a switcher, as in @Ba2023. The dogmatic modeler can be seen as a switcher with an infinitely sticky initial belief
and this is the sense in which the two are nested. Both theories produce different predictions about the agent's behavior. I discuss both 
in what follows.

### The Dogmatic Modeler

A dogmatic agent does not update their beliefs about $\theta$, instead, he holds a degenerate belief that places probability on 
$\hat{\theta}$, which is potentially misspecified. 
In this case, no matter
how much information he gathers against being of type $\hat{\theta}$, he will not update his beliefs about $\theta$. 
Any discrepancies between the 
observed outcomes and his believed type are incorporated using the Bayes rule to update their beliefs about $\omega$.
@Heidhues2018 show that, under certain assumptions on the per-period utility,\footnote{The assumptions are that 
$u$ is twice continuously differentiable with: (i)$u_{ee}<0$ and $u_e(\underline{e} \theta, \omega)>0>u_e(\Bar{e}, \theta, \omega)$, 
(ii) $u_{\theta}, u_{\omega}>0$ and (iii) $u_{e\theta}<0$ and $u_{e\omega}>0$. The direction of the derivatives is a normalization
and the results would hold even when the signs are reversed.} 
a dogmatic modeler will inevitably fall into a self-defeating equilibrium. The equilibrium will be such that
the outcomes they observe reinforce their belief on $\omega$ in such a way that as $t\to\infty$ the agent will 
be sure that
the state is some $\omega^{'}$ consistent with their believed type and the observed data. In other words, 
they will be in a self-defeating equilibrium with a stable belief that places probability one on the incorrect
parameters $(\hat{\theta}, \omega_{\infty})$

The mechanism by which the dogmatic agent falls into the self-defeating equilibrium is the following: Suppose the agent holds the 
misspecified belief that they are type $\hat{\theta}>\theta^*$. For any prior over $\omega$, the agent will be disappointed by the outcome.
He expected a gain of $h(\hat{\theta}, \mathbb{E}(\omega), a)$ but instead observes  $h(\hat{\theta}, \omega^*, a)$. 
There are two 
possible sources for the disappointment, the first is that the realized state is lower than the expected state. The second source is that the
agent is of type $\theta^*$ and therefore, for all possible states, his gain will be lower than what he expected. Because the agent is dogmatic,
he will not update his beliefs about $\theta$ and as a consequence will attribute the disappointment to the state being lower than expected.
He will continue to update in this way until he converges to a belief about $\omega$ that is stable. Such a belief will 
explain the observed utility perfectly and allow the agent to rationalize his dogmatic belief about $\theta$. 
Under the assumptions of @Heidhues2018, there is a unique value of $\omega$ at which the belief is stable, I will refer to such value
as $\omega_\infty$.
This mechanism is further illustrated in Example 1.

@@

_Under this specification, the myopic optimal effort level is $a_t^* = \omega^*$. Nonetheless, because the agent does not know $\omega^*$,
he will choose $a_t = \mathbb{E}_t(\omega)$ where the expectation is taken with respect to the agent's belief at the beginning of
period $t$. If he does not revise his effort choice for $k$ periods, he will receive 
an average utility of 
$(\theta^*+a_t^*)\omega^* - \frac{1}{2}a_t^{*2}$ but he was expecting an average utility of 
$(\hat{\theta}\theta+a_t^*)\omega^* - \frac{1}{2}a_t^{*2}$. In response, he will apply Bayes rule to update his beliefs about
$\omega$ to get the posterior belief with $\mathbb{E}_{t+k}[\omega] = \frac{(\theta^{*} + \omega^{*})\omega^{*}}{\hat{\theta} + \omega^{*}}$
which is lower than the initial belief. This will cause the agent to choose a lower effort at $t+k$. As a result, he will again
receive an average utility that is lower than what he expected which will cause his belief to drift further down. This process will
continue until the average utility equals his expected utility under the dogmatic belief that assigns probability 1 to $\hat{\theta}$. 
At that point, the student will have reached a self-defeating equilibrium and he will continue to choose sub-optimal effort forever. _

Although the model of a dogmatic modeler rationalizes the prevalence of overconfident (underconfident) beliefs, the
assumption that the agent has a degenerate belief and no mechanism through which he can update such belief is very restrictive. An
alternative approach is proposed by @Ba2023. She proposes an extension of the dogmatic agent who is able to jump from one dogmatic 
belief to another. By doing so, the agent might end up being dogmatic and correctly specified.

### The Switcher

An agent is a \emph{switcher} if they behave as a dogmatic, but is willing to entertain the possibility that they are of a different type.
In particular, when they start off as a misspecified dogmatic, they are willing to switch to a different dogmatic belief if the data is
convincing enough. Their prior is still degenerate and assigns probability one to a particular type, and zero to 
all other types. This means that a Bayesian update on $\theta$ does not change their beliefs about the type.
However, they are willing to entertain two such beliefs and have a mechanism by which they decide which belief to adopt at any period $t$.

In order to abandon their initial dogmatic belief, the agent needs to observe a sequence of outcomes 
that are sufficiently unlikely
to have happened if they were of the type they initially believed. In order to evaluate if the evidence is convincing enough, 
they keep track of the likelihood that each of the
possible types generated the data. If the likelihood ratio is sufficiently large, the agent will switch to the alternative
and behave as if they are dogmatic about the new type.

In particular, for an agent that starts with a dogmatic belief that they are of type $\hat{\theta}$ but is willing to consider
the alternative explanation that they are of type $\tilde{\theta}$, the agent will switch to the alternative if:

@@

Where $h^t$ is the history of outcomes up to time $t$ and $\alpha$ is the switching threshold. By keeping track of the 
likelihood ratio, the agent can perform a \emph{Bayesian hypothesis test} and adopt the Dogmatic belief that best fits 
the data.\footnote{In a related problem @Schwarstein2021 proposes a similar updating procedure which relies on the Bayesian 
hypothesis test. However, in their model there is a sender who optimally chooses to propose a model that fits the data}
Notice that if $\alpha \to \infty$, the behavior of the switcher will be indistinguishable from that of the Dogmatic modeler. 
In this sense, the switcher is a generalization of the dogmatic type.

By allowing the agent to keep track of the likelihoods and switching to an alternative type, the switcher can avoid the 
self-confirming equilibrium. However, if the prior belief on $omega$ is sufficiently tight around
a self-defeating equilibrium, the switcher might look identical to the dogmatic even in a case where $\alpha$ is not too large.
This happens because under the agent's prior, the likelihood ratio is unlikely to grow as fast as it is needed to escape 
the self-defeating equilibrium. In such situations, we say that the misspecified belief is persistent.

## Framework 2

As in framework 1, the agent is of some type $\theta^* \in \Theta$ and the state is $\omega \sim F$. 
In this case, the agent chooses an action 
$a\in A$
and observes a binary outcome that is either a success or a failure. Denote the outcome by $o \in {s,f}$. The probability of 
observing a success is increasing in $\theta^*$ and in $\omega$. Whenever the agent observes a success, he gets a payoff $v>0$ and
whenever the outcome is a failure, the payoff is 0.  In addition, the probability of success is such that for each state, 
there is a unique optimal action that maximizes the agent's expected payoff.

I focus on two nested theories that have been widely studied within this framework: Full Bayesian updating and self-serving attribution 
bias. I explain each of these classical models of belief updating in what follows

### The Bayesian

@@

### The Self-Serving Updater

@@

# A Unifying Example
The agent can be of one of 3 types: $\theta \in \{\theta_L, \theta_M, \theta_H\}$  with $\theta_H > \theta_M > \theta_L$.
They face an unknown exogenous success rate
$\omega \in \{\omega_L, \omega_M, \omega_H\}$ with $\omega_H>\omega_M>\omega_L$. Each of the values of $\omega$ is realized with 
equal probability. 
The agent knows the distribution of $\omega$ but not its realized value.
